---
title: "Data Visualization in R"
subtitle: "From data to publication-ready static figures"
author: "Siddhartha Bora · West Virginia University"
date: "May 28, 2026"
date-format: long
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: c/t
    show-slide-number: all
    chalkboard: false
    incremental: false
    smaller: false
    scrollable: true
    code-line-numbers: false
    code-overflow: wrap
    embed-resources: true
    fig-align: center
    fig-width: 8
    fig-height: 4.5
    fig-dpi: 150
# Note: to get a printable PDF, render to revealjs HTML and open with
# `?print-pdf` appended to the URL, then use the browser's print dialog.
# See https://quarto.org/docs/presentations/revealjs/presenting.html#print-to-pdf
execute:
  echo: true
  warning: false
  message: false
  cache: false
  fig-format: png
  dev: ragg_png
knitr:
  opts_chunk:
    dev: "ragg_png"
    fig.retina: 2
---

```{r setup}
#| include: false
# Centralized library loads and theme so every chunk inherits them.
# Install dependencies (one time):
#   install.packages(c("tidyverse", "scales", "patchwork", "colorspace",
#                      "sf", "tigris", "ggspatial", "ggrepel",
#                      "ragg", "here", "gapminder"))

library(here)
# Quarto renders from slides/, so R never sees the project-root .Renviron
# (it only checks the working dir then ~/.Renviron). Load it explicitly so the
# Census API key used by demo 19 comes from the project, not a stale ~/.Renviron.
readRenviron(here(".Renviron"))

library(tidyverse)
library(scales)
library(patchwork)
library(colorspace)
# Note: scale_*_viridis_d() / _c() are built into ggplot2 since 3.0.0,
# so no need to load the viridis package separately.
# sf / tigris / ggspatial are loaded only inside map chunks
# so the deck can render even if those are missing.

# Shared helpers (theme_webinar(), okabe_ito(), life_exp_sample())
source(here("R", "helpers.R"))

# Project-wide theme: set once, inherited by every chunk
theme_set(theme_webinar())

# Reproducibility seed for any rng-touching chunks.
set.seed(42)
```

# Welcome {.center background-color="#1a3a2e"}

## Follow along {background-color="#1a3a2e"}

::: {.columns}
::: {.column width="60%"}
All materials live here:

[**github.com/SiddharthaSBora/r-dataviz-webinar**](https://github.com/SiddharthaSBora/r-dataviz-webinar)

::: {.incremental}
- 📊 Slide deck: `slides/`
- 💻 Live-coding companion: `live-coding/`
- 📋 One-page cheatsheet: `cheatsheet/`
- 🧪 21 standalone demo scripts: `demos/`
:::
:::

::: {.column width="40%"}
*Open the repo now if you want to run code along with the talk.*

`git clone https://github.com/SiddharthaSBora/r-dataviz-webinar`
:::
:::

## What this hour gives you {.smaller}

Three concrete things you'll leave with:

::: {.incremental}
- **A workflow** for going from raw data to a polished static figure in R
- **The reasoning** behind design choices: color, type, layout, accessibility
- **A working knowledge of maps in R**: the part most people find hardest
:::

::: {.fragment .fade-in}
This is a general R dataviz talk. The map content suits anyone doing regional comparisons, and the principles apply generally.
:::

## Two things before we start {.smaller}

::: {.incremental}
- **I'm just another R user.** I use these tools day to day and I'm sharing what works for me.
- **This is opinionated.** There are many ways to do everything here. These are choices, not rules, take what's useful and leave the rest.
:::

## A note on editors {.smaller}

None of this depends on your editor, the code is identical everywhere.

::: {.incremental}
- **RStudio** — what I use, and what you'll see today. The classic R IDE.
- **VS Code** — I use it too; solid with the R extension if you already live there.
- **Positron** — Posit's newer IDE, a VS Code–based successor to RStudio. Many people are making the switch.
:::

::: {.fragment .fade-in}
Pick whatever keeps you productive, everything here runs the same in all three. With that, let's get into it.
:::

# Part 0: How to think about data viz {background-color="#1a3a2e"}

::: aside
This section follows the arc of *R for Data Science (2e)*, chapters 1–11, free at [r4ds.hadley.nz](https://r4ds.hadley.nz/)
:::

## The data science "whole game" {.smaller}

```{=html}
<div style="display:flex; justify-content:center; margin-top:0.5em;">
<svg viewBox="0 0 920 340" xmlns="http://www.w3.org/2000/svg" style="max-width:100%; height:auto; font-family:'Source Sans Pro',sans-serif;">
  <defs>
    <marker id="arrow" markerWidth="10" markerHeight="10" refX="8" refY="3" orient="auto" markerUnits="strokeWidth">
      <path d="M0,0 L8,3 L0,6 Z" fill="#2c5f2d"/>
    </marker>
  </defs>

  <!-- Import -->
  <rect x="20" y="150" width="110" height="44" rx="6" fill="#1a3a2e"/>
  <text x="75" y="177" text-anchor="middle" fill="#fff" font-size="18" font-weight="700">Import</text>

  <!-- Tidy -->
  <rect x="170" y="150" width="110" height="44" rx="6" fill="#2c5f2d"/>
  <text x="225" y="177" text-anchor="middle" fill="#fff" font-size="18" font-weight="700">Tidy</text>

  <!-- Understand box (dashed) containing Transform / Visualize / Model -->
  <rect x="330" y="60" width="300" height="230" rx="10" fill="none" stroke="#97bc62" stroke-width="2" stroke-dasharray="6 5"/>
  <text x="480" y="50" text-anchor="middle" fill="#6a6a6a" font-size="15" font-style="italic">Understand</text>

  <rect x="370" y="80" width="220" height="40" rx="6" fill="#97bc62"/>
  <text x="480" y="106" text-anchor="middle" fill="#1a3a2e" font-size="17" font-weight="700">Transform</text>

  <rect x="370" y="155" width="220" height="40" rx="6" fill="#97bc62"/>
  <text x="480" y="181" text-anchor="middle" fill="#1a3a2e" font-size="17" font-weight="700">Visualize</text>

  <rect x="370" y="230" width="220" height="40" rx="6" fill="#97bc62"/>
  <text x="480" y="256" text-anchor="middle" fill="#1a3a2e" font-size="17" font-weight="700">Model</text>

  <!-- internal loop arrows between the three -->
  <line x1="480" y1="120" x2="480" y2="153" stroke="#2c5f2d" stroke-width="2" marker-end="url(#arrow)"/>
  <line x1="480" y1="195" x2="480" y2="228" stroke="#2c5f2d" stroke-width="2" marker-end="url(#arrow)"/>
  <line x1="430" y1="230" x2="430" y2="122" stroke="#2c5f2d" stroke-width="2" marker-end="url(#arrow)"/>

  <!-- Communicate -->
  <rect x="700" y="150" width="130" height="44" rx="6" fill="#1a3a2e"/>
  <text x="765" y="177" text-anchor="middle" fill="#fff" font-size="18" font-weight="700">Communicate</text>

  <!-- main flow arrows -->
  <line x1="130" y1="172" x2="166" y2="172" stroke="#2c5f2d" stroke-width="2.5" marker-end="url(#arrow)"/>
  <line x1="280" y1="172" x2="326" y2="172" stroke="#2c5f2d" stroke-width="2.5" marker-end="url(#arrow)"/>
  <line x1="634" y1="172" x2="696" y2="172" stroke="#2c5f2d" stroke-width="2.5" marker-end="url(#arrow)"/>
</svg>
</div>
```

::: {.fragment}
Today we live in the middle: **Tidy → Transform → Visualize → Communicate**, the visual half of the cycle.
:::

::: aside
Diagram after Wickham, Çetinkaya-Rundel & Grolemund, *R for Data Science (2e)*, Introduction.
:::

## Why visualize *first*? {.smaller}

::: {.columns}
::: {.column width="55%"}
R4DS teaches `ggplot` in **chapter 1**, before any data wrangling.

Visualization is how you *understand* data, as much as how you present it.

::: {.fragment}
**Anscombe's quartet**, four datasets with nearly identical means, variances, and correlations…
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-height: 4
library(tidyr)

anscombe_tidy <- anscombe |>
  pivot_longer(everything(),
               names_to = c(".value", "set"),
               names_pattern = "(.)(.)") |>
  mutate(set = paste0("Set ", set))

ggplot(anscombe_tidy, aes(x, y)) +
  geom_point(color = "#2c5f2d", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#97bc62", linewidth = 0.8) +
  facet_wrap(~ set) +
  labs(x = NULL, y = NULL) +
  theme(strip.text = element_text(face = "bold"))
```
:::
:::

::: {.fragment}
…yet they look **completely different**. Summary statistics can mislead; pictures don't. **Plot early, plot often.**
:::

## Tidy data: the foundation {.smaller}

::: {.columns}
::: {.column width="55%"}
Three rules *(R4DS Ch. 5)*:

::: {.incremental}
1. Each **variable** is a column
2. Each **observation** is a row
3. Each **value** is a cell
:::

::: {.fragment}
Tidy data is what lets every downstream step (transform, plot, model) work with the same tools.
:::
:::

::: {.column width="45%"}
::: {.fragment}
```{r}
#| echo: false
#| fig-height: 3.6
grid <- expand.grid(col = 1:3, row = 1:4)
ggplot(grid, aes(col, -row)) +
  geom_tile(fill = "#97bc62", color = "white", linewidth = 2) +
  annotate("text", x = 2, y = -0.3, label = "← variables (columns)",
           size = 4, color = "#1a3a2e", fontface = "italic") +
  annotate("text", x = 0.2, y = -2.5, label = "observations\n(rows) ↓",
           size = 4, color = "#1a3a2e", fontface = "italic", hjust = 1) +
  coord_cartesian(xlim = c(-1.2, 3.7), clip = "off") +
  theme_void()
```
:::
:::
:::

> "Tidy datasets are all alike; every messy dataset is messy in its own way." (Wickham)

## Data wrangling: the real work {.smaller}

Before any plot, the data has to be **tidy**. In practice, getting it there is the **biggest challenge**, often most of the work.

::: {.columns}
::: {.column width="55%"}
The everyday verbs *(R4DS Ch. 3 & 5)*:

::: {.incremental}
- **`filter()`** — keep rows
- **`select()`** — keep columns
- **`mutate()`** — add / change columns
- **`group_by()` + `summarise()`** — aggregate
- **`pivot_longer()` / `pivot_wider()`** — reshape
- **`*_join()`** — combine tables
:::
:::

::: {.column width="45%"}
::: {.fragment}
**Real data is messy:**

- Values stored in column names
- One cell holding two variables
- Dates and numbers stored as text
- Missing values, inconsistent labels
- Several tables that must be joined
:::
:::
:::

::: {.fragment}
We won't go deep today, it's a topic of its own. Just know: **if a plot feels hard, the fix is usually upstream, in the data.**
:::

::: aside
A whole subject in itself, see *R for Data Science (2e)*, chapters 3 (transform) and 5 (tidy).
:::

## The grammar of graphics {.smaller}

A plot is built from independent **layers** *(R4DS Ch. 1 & 9)*:

```{=html}
<div style="display:flex; justify-content:center; margin-top:0.3em;">
<svg viewBox="0 0 760 300" xmlns="http://www.w3.org/2000/svg" style="max-width:100%; height:auto; font-family:'Source Sans Pro',sans-serif;">
  <g>
    <rect x="40"  y="40" width="180" height="46" rx="6" fill="#1a3a2e"/>
    <text x="130" y="69" text-anchor="middle" fill="#fff" font-size="17" font-weight="700">data</text>
    <text x="240" y="69" text-anchor="middle" fill="#2c5f2d" font-size="26" font-weight="700">+</text>

    <rect x="270" y="40" width="220" height="46" rx="6" fill="#2c5f2d"/>
    <text x="380" y="69" text-anchor="middle" fill="#fff" font-size="16" font-weight="700">aesthetic mapping</text>
    <text x="510" y="69" text-anchor="middle" fill="#2c5f2d" font-size="26" font-weight="700">+</text>

    <rect x="540" y="40" width="180" height="46" rx="6" fill="#97bc62"/>
    <text x="630" y="69" text-anchor="middle" fill="#1a3a2e" font-size="17" font-weight="700">geom</text>
  </g>

  <text x="130" y="115" text-anchor="middle" fill="#6a6a6a" font-size="13">the table</text>
  <text x="380" y="115" text-anchor="middle" fill="#6a6a6a" font-size="13">columns to visual channels</text>
  <text x="630" y="115" text-anchor="middle" fill="#6a6a6a" font-size="13">how rows are drawn</text>

  <text x="380" y="165" text-anchor="middle" fill="#1a3a2e" font-size="17" font-weight="700">… then optional layers:</text>

  <g font-size="15">
    <rect x="90"  y="190" width="120" height="38" rx="6" fill="#eef3e6" stroke="#97bc62"/>
    <text x="150" y="214" text-anchor="middle" fill="#1a3a2e">scales</text>
    <rect x="230" y="190" width="120" height="38" rx="6" fill="#eef3e6" stroke="#97bc62"/>
    <text x="290" y="214" text-anchor="middle" fill="#1a3a2e">facets</text>
    <rect x="370" y="190" width="120" height="38" rx="6" fill="#eef3e6" stroke="#97bc62"/>
    <text x="430" y="214" text-anchor="middle" fill="#1a3a2e">coords</text>
    <rect x="510" y="190" width="120" height="38" rx="6" fill="#eef3e6" stroke="#97bc62"/>
    <text x="570" y="214" text-anchor="middle" fill="#1a3a2e">theme</text>
  </g>
</svg>
</div>
```

::: {.fragment}
This *is* how `ggplot2` works. Learn the grammar once; every plot is a recombination of these pieces.
:::

## Aesthetic mappings & how we read them {.smaller}

::: {.columns}
::: {.column width="50%"}
**Mapping** = assigning a data variable to a visual channel:

- `x`, `y` → **position**
- `color`, `fill` → hue
- `shape`, `linetype` → form
- `size` → area

*(R4DS Ch. 1 & 9)*
:::

::: {.column width="50%"}
::: {.fragment}
**But we don't read all channels equally well.**

Accuracy of human perception, best → worst:

1. Position along a scale
2. Length
3. Angle / slope
4. Area
5. Color / shade / density
:::
:::
:::

::: {.fragment}
This is *why* bar charts beat pie charts, and why color should carry **categories or emphasis**, not precise quantities.
:::

::: aside
Cleveland & McGill (1984), graphical perception. Echoed throughout R4DS Ch. 9–11.
:::

## Exploratory data analysis (EDA) {.smaller}

EDA is a **mindset**, not a checklist *(R4DS Ch. 10)*:

::: {.incremental}
- Generate questions about your data
- Search for answers by visualizing, transforming, and modeling
- Use what you learn to refine the questions: and repeat
:::

::: {.fragment}
Two questions drive most of it:

- What type of **variation** occurs within my variables?
- What type of **covariation** occurs between my variables?
:::

::: {.fragment}
> "EDA is not a formal process with a strict set of rules… it is a state of mind." (R4DS)

The plots you make *for yourself* (fast, ugly, many) differ from the ones you make *for others* (slow, polished, few).
:::

## Communication {.smaller}

The shift from **exploratory** (for you) to **explanatory** (for others) *(R4DS Ch. 11)*:

::: {.columns}
::: {.column width="50%"}
**Exploratory plot**

- Default labels
- Default colors
- Made in seconds
- Audience: you
:::

::: {.column width="50%"}
::: {.fragment}
**Explanatory plot**

- Informative title + subtitle
- Deliberate color
- Annotations, source caption
- Audience: everyone else
:::
:::
:::

::: {.fragment}
A figure that communicates needs more than the ggplot default. Most of today's hands-on portion is this transformation, **layer by layer**.
:::

## Workflow & reproducibility {.smaller}

R4DS interleaves "workflow" chapters *(Ch. 2, 4, 6, 8)*. The craft around the code matters as much as the code:

::: {.incremental}
- **Projects, not `setwd()`** *(Ch. 6)*: your analysis should run on anyone's machine
- **Readable code style** *(Ch. 4)*: code is read far more often than written
- **Knowing how to get help** *(Ch. 8)*: reprexes, documentation, the community
- **Scripts you can re-run**: reproducibility built in from the start
:::

::: {.fragment}
That's why this webinar ships as an **R project**, not loose scripts.
:::

## Why R + the tidyverse {.smaller}

The handoff to the hands-on portion:

::: {.incremental}
- The **grammar of graphics** is literally how `ggplot2` is built
- The **tidyverse** is one coherent toolkit for the whole game: import, tidy, transform, visualize, communicate
- **Reproducible**: scriptable, version-controllable, re-runnable
- **Publication-quality** static output, and it's **free**
:::

::: {.fragment}
We've covered the *what* and the *why*.

**Now the *how*.**
:::

# Demos 1–2 · Data first {background-color="#1a3a2e"}

## Demo 1: Tidy data {.smaller}

`▶ demos/01-tidy-data.R`

```{r}
#| echo: false
#| fig-height: 2.2

library(tibble)

untidy <- tibble(
  student = c(1, 2, 3),
  english = c(70, 78, 82),
  math    = c(87, 65, 91),
  science = c(100, 83, 88)
)

tidy <- tibble(
  student = rep(1:3, each = 3),
  subject = rep(c("english","math","science"), 3),
  score   = c(70,87,100, 78,65,83, 82,91,88)
)
```

::: {.columns}
::: {.column width="48%"}
**Untidy**, subjects in columns (3 rows)
```{r}
#| echo: false
knitr::kable(untidy)
```
:::

::: {.column width="48%"}
**Tidy**, one row per (student, subject) (9 rows)
```{r}
#| echo: false
knitr::kable(tidy)
```
:::
:::

::: {.fragment}
**The rules** (Wickham 2014):

1. Each **variable** is a column
2. Each **observation** is a row
3. Each **value** is a cell
:::

::: {.fragment .fade-in}
**Why we care:** `ggplot2` is *designed* for tidy data. Long form unlocks aesthetic mappings like `color = subject`.
:::

## Demo 2: Long vs wide {.smaller}

`▶ demos/02-long-vs-wide.R`

The same data, two shapes:

```{r}
library(tidyr)

scores_wide <- tibble(
  student = c(1, 2, 3),
  english = c(70, 78, 82),
  math    = c(87, 65, 91),
  science = c(100, 83, 88)
)

# WIDE -> LONG
scores_long <- scores_wide |>
  pivot_longer(
    cols      = -student,
    names_to  = "subject",
    values_to = "score"
  )

scores_long
```

## The reverse direction

```{r}
# LONG -> WIDE
scores_long |>
  pivot_wider(names_from = subject, values_from = score)
```

**Rule of thumb:**

- Storing, analyzing, plotting → **long**
- Final summary table for a human reader → **wide**

> "If your plot feels hard, your data is probably wide."

# Demo 3 · The grammar of graphics {background-color="#1a3a2e"}

## In one sentence

`▶ demos/03-first-ggplot.R`

> Map data variables to visual properties (aesthetics) and pick a geometric object (geom) to represent each row.

::: {.fragment}
Every ggplot has three things:
:::

::: {.fragment}
``` r
ggplot(data = <DATA>) +
  aes(x = <VAR>, y = <VAR>, color = <VAR>) +
  geom_<TYPE>()
```
:::

::: {.fragment}
- **`data`**: a data frame
- **`aes()`**: mapping from variables to visual properties
- **`geom_*()`**: the visual representation (line, point, bar, polygon...)
:::

## Same data, different geom {.smaller}

```{r}
#| fig-height: 3.5
p_base <- ggplot(economics, aes(x = date, y = unemploy / 1000))

p_line  <- p_base + geom_line()  + labs(title = "geom_line",  x = NULL, y = "Million")
p_area  <- p_base + geom_area(fill = "#2c5f2d", alpha = 0.6) + labs(title = "geom_area",  x = NULL, y = "Million")
p_point <- p_base + geom_point(alpha = 0.3, size = 0.8) + labs(title = "geom_point", x = NULL, y = "Million")
p_step  <- p_base + geom_step() + labs(title = "geom_step",  x = NULL, y = "Million")

(p_line | p_area) / (p_point | p_step)
```

**The data didn't move. The geom is a *choice* about how to represent it.**

# Demo 4 · A worked example {background-color="#1a3a2e"}

## The plan

`▶ demos/04-layered-improvement.R`

We'll take **one** plot from "default ggplot" to "publication ready" by adding **one layer at a time.**

::: {.fragment}
Sample data: life expectancy at birth in three countries, 1990–2020.
:::

```{r}
life_exp <- life_exp_sample()
life_exp
```

## Step 1: the crudest possible plot

```{r}
#| fig-height: 3.5
p1 <- life_exp |>
  ggplot(aes(x = year, y = life_exp, color = country)) +
  geom_line()

p1
```

::: {.fragment}
Works. Looks bad. **That's the point**, every default ggplot looks like this.
:::

## Step 2: treat the plot as an object

The single most powerful idea in ggplot:

::: {.fragment}
```{r}
#| fig-height: 3
p2 <- p1 +
  labs(
    title    = "Life expectancy at birth, 1990–2020",
    subtitle = "Brazil, India, United States",
    x        = NULL,
    y        = "Years",
    color    = NULL,
    caption  = "Source: illustrative data"
  )

p2
```
:::

::: {.fragment}
**Each `+` is a layer.** You can save plots as objects, add to them later, and reuse them.
:::

## Step 3: thicker lines, end-point markers

```{r}
#| fig-height: 3.5
# Note: linewidth, not size: ggplot 3.4.0 (2022) deprecated `size` for lines.
p3 <- p2 +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5)

p3
```

## Step 4: pick a deliberate color palette

```{r}
#| fig-height: 3.5
# Viridis: perceptually uniform AND colorblind-safe.
p4 <- p3 +
  scale_color_viridis_d(option = "D", end = 0.85)

p4
```

::: {.fragment}
We'll come back to color choices in a moment.
:::

## Step 5: apply a clean theme

```{r}
#| fig-height: 3.5
p5 <- p4 +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

p5
```

## Step 6: fix axis breaks and limits

```{r}
#| fig-height: 3.5
p6 <- p5 +
  scale_x_continuous(breaks = seq(1990, 2020, 10)) +
  scale_y_continuous(limits = c(55, 80), labels = scales::label_number(suffix = " yr"))

p6
```

## Before and after {.smaller}

```{r}
#| echo: false
#| fig-height: 3.5
p1_simple <- life_exp |>
  ggplot(aes(x = year, y = life_exp, color = country)) +
  geom_line() +
  labs(title = "Step 1")

p_final <- p6 + labs(title = "Step 6")

p1_simple + p_final
```

**Six small, named layers. Each one does one thing.**

> You didn't write a 50-line plotting function. You assembled a figure.

# Demo 5 · Color, done deliberately {background-color="#1a3a2e"}

## Three kinds of color scale {.smaller}

`▶ demos/05-color-palettes.R`

```{r}
#| echo: false
#| fig-height: 4

library(scales)

# Three demonstration data sets
df_qual <- tibble(x = letters[1:6], y = c(3, 5, 2, 7, 4, 6))
df_seq  <- tibble(x = 1:10, y = 1:10)
df_div  <- tibble(x = -5:5, y = -5:5)

p_qual <- ggplot(df_qual, aes(x, y, fill = x)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Qualitative", subtitle = "unordered categories, Set2") +
  theme(legend.position = "none")

p_seq <- ggplot(df_seq, aes(x, y = 1, fill = y)) +
  geom_tile() +
  geom_text(aes(label = y), color = "white", size = 3) +
  scale_fill_viridis_c() +
  labs(title = "Sequential", subtitle = "ordered, low → high, viridis") +
  theme(legend.position = "none", axis.text.y = element_blank(),
        axis.title.y = element_blank())

p_div <- ggplot(df_div, aes(x, y = 1, fill = x)) +
  geom_tile() +
  geom_text(aes(label = x), color = "grey20", size = 3) +
  scale_fill_distiller(palette = "RdBu", limits = c(-5, 5)) +
  labs(title = "Diverging", subtitle = "symmetric around midpoint, RdBu") +
  theme(legend.position = "none", axis.text.y = element_blank(),
        axis.title.y = element_blank())

p_qual / p_seq / p_div
```

**First pick the *type*, then the palette.** Diverging only when your data has a meaningful midpoint (zero, mean, baseline).

## The palette functions you'll actually use

| Type | Discrete | Continuous |
|------|----------|------------|
| Qualitative | `scale_*_brewer(palette = "Set2")` | — |
| Qualitative (colorblind-safe) | `scale_*_viridis_d()`, Okabe-Ito | — |
| Sequential | `scale_*_viridis_d()` | `scale_*_viridis_c()` |
| Sequential (alt) | `scale_*_distiller(palette = "Blues")` | `scale_*_gradient()` |
| Diverging | `scale_*_distiller(palette = "RdBu")` | `scale_*_gradient2()` |
| Manual | `scale_*_manual(values = c(...))` | `scale_*_gradientn(colours = c(...))` |

Replace `*` with `color` or `fill` depending on the aesthetic.

## Default ggplot colors aren't a great default {.smaller}

```{r}
#| echo: false
#| fig-height: 3.5
df <- tibble(g = LETTERS[1:5], y = c(4, 7, 3, 5, 6))

p_default <- ggplot(df, aes(g, y, fill = g)) + geom_col() +
  labs(title = "Default (hue_pal)") + theme(legend.position = "none")

p_viridis <- ggplot(df, aes(g, y, fill = g)) + geom_col() +
  scale_fill_viridis_d(end = 0.9) +
  labs(title = "Viridis (perceptually uniform)") + theme(legend.position = "none")

p_okabe <- ggplot(df, aes(g, y, fill = g)) + geom_col() +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2")) +
  labs(title = "Okabe-Ito (colorblind-safe)") + theme(legend.position = "none")

p_default + p_viridis + p_okabe
```

::: {.fragment}
Defaults are *colorful*. They are not *colorblind-safe*, not *perceptually uniform*, and not *meaningful*.
:::

::: {.fragment}
**Pick a palette deliberately.** Once per project, not once per plot.
:::

# Demo 6 · Type, theme, polish {background-color="#1a3a2e"}

## Built-in themes {.smaller}

`▶ demos/06-themes-and-fonts.R`

```{r}
#| echo: false
#| fig-height: 3.5

p_base_demo <- ggplot(life_exp |> filter(country == "Brazil"),
                     aes(year, life_exp)) +
  geom_line(linewidth = 1, color = "#2c5f2d") +
  labs(title = "Brazil")

(p_base_demo + theme_gray()    + ggtitle("theme_gray (default)")) +
(p_base_demo + theme_minimal() + ggtitle("theme_minimal")) +
(p_base_demo + theme_bw()      + ggtitle("theme_bw")) +
(p_base_demo + theme_classic() + ggtitle("theme_classic"))
```

`theme_minimal()` and `theme_classic()` are safe defaults for most published work.

## The `theme()` function: what you actually customize

```{r}
#| eval: false
theme(
  # Titles and captions
  plot.title    = element_text(face = "bold", size = 14),
  plot.subtitle = element_text(color = "grey40"),
  plot.caption  = element_text(color = "grey50", hjust = 0),

  # Axes
  axis.title = element_text(size = 12),
  axis.text  = element_text(size = 10),

  # Legend
  legend.position = "bottom",       # "top", "right", "none", or c(x, y)
  legend.title    = element_blank(),

  # Panel
  panel.grid.minor = element_blank(),
  panel.grid.major.x = element_blank()
)
```

::: aside
**Tip:** Set a project-wide theme with `theme_set(theme_minimal() + theme(...))` at the top of your script. Then every plot inherits it.
:::

## Format numbers properly: the `scales` package

```{r}
#| fig-height: 3
library(scales)

ggplot(economics, aes(date, unemploy * 1000)) +
  geom_line(linewidth = 1, color = "#2c5f2d") +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  labs(title = "US unemployed persons", x = NULL, y = NULL)
```

::: {.fragment}
Useful: `label_number()`, `label_percent()`, `label_comma()`, `label_dollar()`, `label_date()`
:::

## Fonts: and the export gotcha

::: {.incremental}
- Fonts that look fine on screen often vanish or substitute in PDF export
- The fix: use the **`ragg`** graphics device for PNG, **`cairo_pdf`** for PDF
- Old advice (`type = "cairo"` in `ggsave`) no longer works: `ragg` replaced it
:::

::: {.fragment}
More in the export section. The device matters.
:::

# Demo 7 · Faceting {background-color="#1a3a2e"}

## Small multiples {.smaller}

`▶ demos/07-faceting.R`

```{r}
#| fig-height: 3.5
# Use gapminder if available (more interesting), else fall back to life_exp.
if (requireNamespace("gapminder", quietly = TRUE)) {
  gapminder::gapminder |>
    filter(continent == "Americas",
           country %in% c("Brazil", "Argentina", "Mexico", "Chile",
                          "Colombia", "Peru")) |>
    ggplot(aes(year, lifeExp)) +
    geom_line(linewidth = 1, color = "#2c5f2d") +
    facet_wrap(~ country) +
    labs(title = "Life expectancy, selected Americas",
         x = NULL, y = "Years")
} else {
  life_exp |>
    ggplot(aes(year, life_exp)) +
    geom_line(linewidth = 1, color = "#2c5f2d") +
    geom_point(size = 2, color = "#2c5f2d") +
    facet_wrap(~ country) +
    labs(title = "Life expectancy by country", x = NULL, y = "Years")
}
```

**One crowded plot becomes six clear panels.**

## `facet_wrap` vs `facet_grid` {.smaller}

```{r}
#| eval: false
# Wrap one variable into a grid of panels
ggplot(...) + facet_wrap(~ region)

# Cross two variables: rows by year, columns by region
ggplot(...) + facet_grid(year ~ region)

# Free scales: let each panel pick its own y range
ggplot(...) + facet_wrap(~ region, scales = "free_y")

# Add row/column labels via labeller
ggplot(...) + facet_wrap(~ region, labeller = label_both)
```

> **When in doubt, facet.** Five small clear panels beat one cluttered plot.

# Demo 8 · Accessibility {background-color="#1a3a2e"}

## Why this matters

`▶ demos/08-accessibility.R`

::: {.incremental}
- **Federal reports** (USDA, NIH, EPA) require Section 508 / ADA compliance
- **Journals** increasingly require accessibility statements
- Accessibility-first design produces clearer figures **for everyone**
:::

::: {.fragment}
This isn't a checklist. It's better design.
:::

## What a colorblind reader sees {.smaller}

```{r}
#| echo: false
#| fig-height: 3.5
library(colorspace)

# Build a bad-on-purpose plot using default ggplot colors
df <- tibble(
  x = rep(1:10, 4),
  y = c(1:10, 2:11, 3:12, 4:13),
  g = rep(c("Apples", "Pears", "Plums", "Grapes"), each = 10)
)
bad_p <- ggplot(df, aes(x, y, color = g)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("red", "darkred", "orange", "green")) +
  labs(title = "Original") +
  theme(legend.position = "right")

# Same plot through deuteranope simulation
bad_p_deutan <- ggplot(df, aes(x, y, color = g)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = colorspace::deutan(c("red", "darkred", "orange", "green"))) +
  labs(title = "Deuteranopia simulation") +
  theme(legend.position = "right")

bad_p + bad_p_deutan
```

The categories that were distinct become indistinguishable.

## Rule 1: don't encode meaning in color alone

```{r}
#| fig-height: 3
# Color + linetype: redundant encoding survives grayscale & colorblindness
ggplot(life_exp, aes(year, life_exp,
                     color = country,
                     linetype = country)) +
  geom_line(linewidth = 1.1) +
  geom_point(aes(shape = country), size = 2.5) +
  scale_color_viridis_d(end = 0.85) +
  labs(title = "Color + linetype + shape, three ways to tell them apart",
       x = NULL, y = "Years")
```

::: {.fragment}
**The test:** print your figure in grayscale. Can you still read it?
:::

## Rule 2: test what you ship

```{r}
#| eval: false
# colorspace package: installed automatically with ggplot2 via the
# scales package, so you almost certainly have it already.
library(colorspace)

# Simulate any of three common color vision deficiencies on your plot:
deutan(c("red", "green", "blue"))   # deuteranopia (red-green)
protan(c("red", "green", "blue"))   # protanopia (red-green)
tritan(c("red", "green", "blue"))   # tritanopia (blue-yellow)

# For an interactive emulator on a full plot:
colorspace::cvd_emulator()
```

```{r}
#| eval: false
# colorBlindness package: 6-panel side-by-side simulation
library(colorBlindness)
cvdPlot(my_plot)   # renders original + 5 deficiency simulations
```

## Rule 3: alt text + table fallbacks (Quarto)

For HTML / web output, add a `fig-alt` chunk option to every figure:

````markdown
---
title: "My report"
format: html
---

```r
#| label: fig-life-exp
#| fig-cap: "Life expectancy in three countries, 1990–2020"
#| fig-alt: "Line chart showing life expectancy rising steadily
#|   in Brazil and India but plateauing in the US between
#|   1990 and 2020."
ggplot(life_exp, ...) + ...
```
````

::: {.fragment}
**Alt text describes the takeaway, not the chart type.**

A screen-reader user needs the *insight*, not "a line chart with three lines."
:::

## Accessibility quick checklist {.smaller}

::: {.columns}
::: {.column}
**Color**

- Use viridis or Okabe-Ito by default
- Test with `colorspace::deutan()`
- Never encode meaning in color alone
:::

::: {.column}
**Text**

- ≥10pt captions, ≥12pt body, ≥14pt for projection
- Dark text on light bg (or vice versa)
- Avoid rotated text where possible
:::
:::

::: {.columns}
::: {.column}
**Encoding**

- Color + shape, color + linetype
- Direct labels with `geom_text_repel()`
- Provide data table as fallback
:::

::: {.column}
**Output**

- Alt text on every figure (`fig-alt`)
- Tagged PDFs when possible
- Captions are not alt text
:::
:::

> **Color is supplemental. If your figure dies in grayscale, redesign it.**

# Demos 9–13, 18–19 · Maps in R {background-color="#1a3a2e"}

## The mental model

`▶ demos/09-map-basics-sf.R`

A map is **a data frame** where each row has a **`geometry`** column.

::: {.fragment}
Once you accept that, the rest is just ggplot:
:::

::: {.fragment}
```{r}
#| fig-height: 3.2
library(sf)
# sf ships an example shapefile (North Carolina counties) — no download
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

ggplot(nc) +                     # nc is a data frame...
  geom_sf(aes(fill = BIR74)) +   # ...with a geometry column
  scale_fill_viridis_c()
```
:::

::: {.fragment}
That's it. Everything else is detail.
:::

## Where geometry comes from

```{r}
#| eval: false
library(sf)       # the spatial foundation
library(tigris)   # US Census shapefiles
# library(tidycensus)    # for data + geometry in one call
# library(rnaturalearth) # world countries / states

options(tigris_use_cache = TRUE)  # cache shapefiles, saves enormous time

# US states
us_states <- tigris::states(cb = TRUE, year = 2024)

# US counties for a single state
wv_counties <- tigris::counties(state = "WV", cb = TRUE, year = 2024)

# World countries
# world <- rnaturalearth::ne_countries(returnclass = "sf")
```

::: {.fragment}
**`cb = TRUE`** returns simplified "cartographic boundary" files, smaller, faster, fine for choropleths. Use full TIGER lines (`cb = FALSE`) only when you need precise boundaries.
:::

::: aside
The `year` argument controls the Census vintage. Omit it to use the latest available (currently 2024).
:::

## Your first choropleth {.smaller}

`▶ demos/10-map-choropleth.R`

```{r}
#| fig-height: 3.6
#| output-location: column
library(sf)
library(tigris)
library(dplyr)
options(tigris_use_cache = TRUE)

# 1. Get the geometry
us_states <- states(cb = TRUE, year = 2024, resolution = "20m")

# 2. Make some data to join (here: random values; in practice your CSV)
my_data <- tibble(
  state_abbr = state.abb,
  value      = runif(50, 0, 100)
)

# 3. Join data to geometry
states_with_data <- us_states |>
  left_join(my_data, by = c("STUSPS" = "state_abbr"))

# 4. Plot
ggplot(states_with_data) +
  geom_sf(aes(fill = value), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "D") +
  labs(title = "US states, example choropleth", fill = "Value") +
  theme_void()
```

## Real census data: WV counties with `tidycensus` {.smaller}

`▶ demos/19-census-wv-counties.R`

```{r}
#| fig-height: 3.4
#| output-location: column
library(tidycensus)   # data + geometry from the Census API in one call
library(ggplot2)

# Key comes from .Renviron (CENSUS_API_KEY); free signup at
# https://api.census.gov/data/key_signup.html
census_api_key(Sys.getenv("CENSUS_API_KEY"))

# B19013_001 = median household income (ACS 5-year)
wv_income <- get_acs(
  geography = "county",
  state     = "WV",
  variables = "B19013_001",
  year      = 2022,
  survey    = "acs5",
  geometry  = TRUE              # returns an sf data frame
)

ggplot(wv_income) +
  geom_sf(aes(fill = estimate), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "D", labels = scales::label_dollar(scale = 1e-3, suffix = "k")
  ) +
  labs(
    title    = "Median household income, West Virginia counties",
    subtitle = "ACS 5-year estimates, 2018–2022",
    fill     = NULL,
    caption  = "Source: US Census Bureau, ACS via tidycensus"
  ) +
  theme_void(base_size = 12)
```

::: {.fragment}
**`tidycensus::get_acs(geometry = TRUE)`** is the one call you usually want, it joins ACS data to TIGER geometry for you, no manual `left_join()` needed.
:::

## Projections: the one that bites everyone

`▶ demos/11-map-projections.R`

```{r}
library(sf); library(tigris); library(dplyr)
options(tigris_use_cache = TRUE)
us_states <- states(cb = TRUE, year = 2024, resolution = "20m")
conus <- us_states |>                  # drop AK/HI/territories for the comparison
  filter(!STUSPS %in% c("AK", "HI", "PR", "GU", "VI", "MP", "AS"))
```

::: {.columns}
::: {.column width="33%"}
**Default (lon/lat / Web Mercator)**

Stretched. Areas not proportional.

```{r}
#| fig-height: 3
ggplot(conus) +
  geom_sf(fill = "#97bc62", color = "white", linewidth = 0.2)
```
:::

::: {.column width="33%"}
**Albers Equal Area (EPSG:5070)**

The right choice for continental US.

```{r}
#| fig-height: 3
ggplot(conus) +
  geom_sf(fill = "#97bc62", color = "white", linewidth = 0.2) +
  coord_sf(crs = "EPSG:5070")
```
:::

::: {.column width="33%"}
**With Alaska + Hawaii rescaled**

`shift_geometry()` rescales them so the whole country fits.

```{r}
#| fig-height: 3
us_states |>
  tigris::shift_geometry() |>
  ggplot() +
  geom_sf(fill = "#97bc62", color = "white", linewidth = 0.2) +
  coord_sf(crs = "EPSG:5070")
```
:::
:::

::: {.fragment}
- Continental US → **EPSG:5070** (Albers Equal Area)
- World maps → **EPSG:8857** (Equal Earth) or **"+proj=robin"** (Robinson)
- Picking a projection for somewhere else → `crsuggest::suggest_crs()`
:::

## Faceted comparison map

`▶ demos/12-map-faceted-comparison.R`

```{r}
#| fig-height: 3.4
#| output-location: column
# Compare two regions side by side: the "small multiples for maps" pattern
library(sf); library(tigris); library(dplyr); library(scales)
options(tigris_use_cache = TRUE)
set.seed(42)
two_states <- counties(state = c("WV", "OH"), cb = TRUE, year = 2024) |>
  mutate(state_name = if_else(STATEFP == "39", "Ohio", "West Virginia"),
         value      = runif(n(), 5, 25))    # demo data; your CSV in practice

ggplot(two_states) +
  geom_sf(aes(fill = value), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "D", labels = label_number(suffix = "%")) +
  facet_wrap(~ state_name) +                # coord_sf() forbids scales = "free"
  labs(title    = "County-level indicator comparison",
       subtitle = "Ohio vs West Virginia, demo data",
       fill     = NULL) +
  theme_void(base_size = 13) +
  theme(strip.text = element_text(face = "bold", size = 14))
```

**`facet_wrap` works on maps the same way it works on regular plots.**

## Map polish: what separates competent from great {.smaller}

`▶ demos/13-map-polish.R`

```{r}
#| fig-height: 3.6
#| output-location: column
library(sf); library(tigris); library(dplyr); library(scales)
library(ggspatial)   # scale bars and north arrows
options(tigris_use_cache = TRUE)
set.seed(42)
wv <- counties(state = "WV", cb = TRUE, year = 2024) |>
  mutate(deviation = runif(n(), -10, 10))    # deviation from state average

ggplot(wv) +
  geom_sf(aes(fill = deviation), color = "white", linewidth = 0.2) +
  # 1. Right scale type: diverging, because the data has a midpoint (0)
  scale_fill_distiller(palette = "RdBu", direction = -1,
                       limits = c(-10, 10), labels = label_number(suffix = "pp")) +
  # 2. Scale bar + north arrow, only when geographic precision matters
  annotation_scale(location = "br", bar_cols = c("grey20", "white")) +
  annotation_north_arrow(location = "tr", style = north_arrow_minimal()) +
  # 3. theme_void() drops axis chrome; 4. always caption the source
  labs(title   = "West Virginia counties: deviation from state average",
       caption = "Source: simulated data; geometry from US Census TIGER/Line") +
  theme_void(base_size = 13) +
  theme(plot.title   = element_text(face = "bold"),
        plot.caption = element_text(color = "grey50", hjust = 0))
```

::: {.fragment}
> If you can make a ggplot, you can make a map. The data has a geometry column; everything else is the same.
:::

## Rasters: when your data is a grid {.smaller}

`▶ demos/18-raster-basics.R`

Vectors (points / lines / polygons) cover most thematic maps. **Rasters** are the other half: a regular grid of cells, each holding a value. Think satellite imagery, elevation (DEMs), land cover, temperature, precipitation.

```{r}
#| fig-height: 3.2
#| output-location: column
library(terra)       # modern raster engine (successor to {raster})
library(tidyterra)   # ggplot geoms for SpatRaster

# terra ships an example DEM (elevation, Luxembourg) — no download
dem <- rast(system.file("ex/elev.tif", package = "terra"))

ggplot() +
  geom_spatraster(data = dem) +
  scale_fill_viridis_c(name = "Elevation (m)", na.value = NA) +
  labs(title = "Elevation: terra example DEM") +
  theme_minimal()
```

::: {.fragment}
- **`terra`**: read, crop, mask, reproject, algebra; replaces older `raster` package
- **`stars`**: multi-dimensional raster cubes (time, bands); plays well with `sf`
- **`tidyterra`**: `geom_spatraster()` so rasters live inside the same ggplot grammar
:::

::: {.fragment}
> Same rules as vector maps, pick a perceptual scale, project sensibly, label the units.
:::

# Demo 14 · Combining plots {background-color="#1a3a2e"}

## patchwork: `p1 + p2` is all you need to know

`▶ demos/14-patchwork.R`

```{r}
#| fig-height: 3
p_left <- ggplot(life_exp, aes(year, life_exp, color = country)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(end = 0.85) +
  labs(title = "Trend", x = NULL, y = "Years")

p_right <- ggplot(life_exp |> filter(year == 2020),
                  aes(reorder(country, life_exp), life_exp, fill = country)) +
  geom_col() +
  scale_fill_viridis_d(end = 0.85, guide = "none") +
  labs(title = "2020 snapshot", x = NULL, y = "Years")

p_left + p_right
```

## More patchwork operators

```{r}
#| eval: false
p1 + p2                                  # side by side
p1 / p2                                  # stacked
(p1 + p2) / p3                           # nesting
p1 | p2                                  # also side by side

# Shared annotation
(p1 + p2) +
  plot_annotation(
    title    = "Shared title",
    subtitle = "Across both panels",
    tag_levels = "A"                     # auto-label panels A, B, C...
  )

# Collect legends (so you only show one)
(p1 + p2) + plot_layout(guides = "collect")
```

::: {.fragment}
You'll see `gridExtra::grid.arrange()` in older code. **`patchwork` is the modern default.**
:::

# Demo 15 · Exporting publication-quality output {background-color="#1a3a2e"}

## `ggsave`: the arguments that matter

`▶ demos/15-ggsave-export.R`

```{r}
#| eval: false
ggsave(
  filename = "fig-1.png",
  plot     = p,
  width    = 7,      # build at FINAL size, don't rescale later
  height   = 4,
  units    = "in",
  dpi      = 300,    # 300 for print, 150 for screen
  device   = ragg::agg_png   # ragg handles fonts correctly
)

# For PDF: use cairo_pdf to embed fonts properly
ggsave("fig-1.pdf", plot = p,
       width = 7, height = 4, units = "in",
       device = cairo_pdf)

# For vector that scales infinitely (web, illustrator)
ggsave("fig-1.svg", plot = p, width = 7, height = 4, units = "in")
```

## Export pitfalls

::: {.incremental}
- **Resizing in Word/PowerPoint after export**: text gets warped. Set the size in `ggsave`.
- **Fonts substituting in PDF**: use `device = cairo_pdf`
- **PNG blurry when zoomed**: use SVG or PDF for vector
- **"Looks fine on my laptop, terrible on the projector"**: always test
- **Different output on different OS**: `ragg` minimizes this; old `png()` doesn't
:::

::: {.fragment}
> **Build the figure at its final size. Don't rescale.**
:::

# Demo 16 · Common pitfalls {background-color="#1a3a2e"}

## Misleading visualizations to avoid {.smaller}

`▶ demos/16-common-pitfalls.R`

::: {.columns}
::: {.column}
- **Truncated y-axis on bar charts**: bars no longer represent magnitude
- **Dual y-axes**: almost always misleading; one of them is a lie
- **Pie charts beyond 3 slices**: humans can't compare angles
- **3D anything**: perspective distorts area
:::

::: {.column}
- **Rainbow / jet colormaps**: perceptually non-uniform; fail under colorblindness
- **Encoding ordering in hue alone**: luminance carries ordering, hue doesn't
- **More gridlines than data points**: declutter
- **Tiny axis labels on big figures**: readable at *final* size
:::
:::

## Portability and reproducibility pitfalls

::: {.incremental}
- **Hardcoded file paths** (`/Users/me/...`) → use `here::here("data", "file.csv")`
- **`setwd()` anywhere in code** → use RStudio projects
- **"It worked yesterday"** → `renv::snapshot()` locks package versions
- **Random seeds not set** → `set.seed(42)` whenever randomness is involved
- **No `sessionInfo()`** → put it at the end of every analysis file
:::

> **Future-you and reviewer-you are different people. Code for both.**

# Demo 17 · Reproducibility {background-color="#1a3a2e"}

## Make it run on someone else's laptop

`▶ demos/17-reproducibility.R`

::: {.incremental}
- **RStudio projects**: never `setwd()`; let the project anchor everything
- **`here::here("data", "file.csv")`**: paths that work regardless of where the script runs
- **`renv`**: lockfile of every package version, like Python's `requirements.txt`
- **`sessionInfo()`** at the end of every notebook
- **Git**: even just `git init` is better than nothing
- **A README**: what this project does, how to install, how to run
:::

::: {.fragment}
> **If you can't hand your project to a stranger and have it run in 10 minutes, it's not done.**
:::

## A minimal reproducible setup

```{r}
#| eval: false
# At the top of every analysis script:
library(here)
library(tidyverse)
library(sf)

set.seed(42)

# Use here() for every path
data <- read_csv(here("data", "raw", "my-data.csv"))

# At the bottom:
sessionInfo()
```

```{r}
#| eval: false
# In the R console, once per project:
renv::init()        # set up renv
renv::snapshot()    # lock current package versions to renv.lock

# On another machine, to install the same versions:
renv::restore()
```

# Demo 20 · Beyond static {background-color="#1a3a2e"}

## When static isn't enough

`▶ demos/20-interactive-plotly.R`

```{r}
#| eval: false
# Turn any ggplot into an interactive HTML widget in one line
library(plotly)
ggplotly(p)
```

```{r}
#| eval: false
# Interactive web maps (tile-based, like Google Maps)
library(leaflet)
leaflet(my_sf_data) |>
  addTiles() |>
  addPolygons(fillColor = ~viridis::viridis(10)[as.numeric(cut(value, 10))])
```

::: {.fragment}
- **`plotly::ggplotly()`**: quickest path to interactive
- **`leaflet`**: interactive maps with tile basemaps
- **`shiny`**: full apps where readers filter / parameterize
- **`tmap`**: alternative mapping framework (v4 in 2025; "thematic maps", switches between static and interactive)
:::

::: {.fragment}
**Caveat:** interactive ≠ better. For a printed report, static is correct.
:::

# Demo 21 · Capstone {background-color="#1a3a2e"}

## Putting it all together

`▶ demos/21-infographic.R`

Every demo taught one move. The last step is **assembly**, chaining them into a single figure that stands on its own.

::: {.fragment}
```{r}
#| eval: false
# One tidy table -> three complementary views, composed with patchwork
figure <- (p_trend) / (p_rank | p_map) +
  plot_layout(heights = c(1.4, 1)) +
  plot_annotation(
    title    = "Two decades of gains across three states",
    subtitle = "A worked example: one tidy table, three views",
    caption  = "Built in R with ggplot2 + patchwork"
  )

ggsave(here("figures", "capstone-infographic.png"),
       figure, device = ragg::agg_png,
       width = 9, height = 7.5, units = "in", dpi = 300)
```
:::

::: {.fragment}
The pipeline, end to end: **tidy data** → a **trend** + a **ranked bar** + a **locator map** → composed, **branded**, and **exported** at a fixed size and DPI so it looks identical everywhere.
:::

::: {.fragment}
**The point:** communication is just assembly. The skill is choosing which views answer the reader's question, then making the result *stand alone*.
:::

# Wrap-up {background-color="#1a3a2e"}

## Useful R packages: a quick tour {.smaller}

::: {.columns}
::: {.column}
**Core plotting**

- **`ggplot2`**: the grammar-of-graphics plotting system
- **`scales`**: axis formatting, transformations, palettes
- **`patchwork`**: combine plots with `+` and `/`
- **`ggrepel`**: labels that don't overlap

**Color**

- **`viridis`**: perceptually-uniform palettes
- **`colorspace`**: color manipulation + CVD simulation
- **`RColorBrewer`**: the ColorBrewer palettes
:::

::: {.column}
**Spatial / maps**

- **`sf`**: vector spatial data as tidy data frames
- **`tigris`**: US Census TIGER shapefiles
- **`tidycensus`**: Census/ACS data with geometry attached
- **`terra`** + **`tidyterra`**: rasters, in ggplot
- **`ggspatial`**: scale bars, north arrows

**Output & interactivity**

- **`ragg`**: fast, high-quality graphics device
- **`plotly`**: turn a ggplot interactive in one line
- **`gt`**: publication-quality tables
:::
:::

::: {.fragment}
**Workflow:** **`here`** (sane paths) · **`renv`** (lock package versions) · **`quarto`** (reproducible documents)
:::

## Resources & books {.smaller}

::: {.columns}
::: {.column}
**Books (all free online)**

- [R for Data Science (2e)](https://r4ds.hadley.nz/): Wickham, Çetinkaya-Rundel, Grolemund
- [ggplot2: Elegant Graphics for Data Analysis (3e)](https://ggplot2-book.org/), Wickham et al.
- [Modern Data Visualization with R](https://rkabacoff.github.io/datavis/), Kabacoff
- [Geocomputation with R](https://r.geocompx.org/): Lovelace, Nowosad, Muenchow
- [Analyzing US Census Data](https://walker-data.com/census-r/), Walker
- [Spatial Data Science (sf/stars)](https://r-spatial.org/book/), Pebesma, Bivand
:::

::: {.column}
**Galleries & inspiration**

- [R Graph Gallery](https://r-graph-gallery.com/)
- [ggplot2 extensions gallery](https://exts.ggplot2.tidyverse.org/gallery/)
- [#TidyTuesday](https://github.com/rfordatascience/tidytuesday), weekly community challenge

**Accessibility**

- [Color Universal Design (Okabe-Ito)](https://jfly.uni-koeln.de/color/)
- [WCAG 2.1 quick reference](https://www.w3.org/WAI/WCAG21/quickref/)

**Quarto**

- [quarto.org/docs/guide](https://quarto.org/docs/guide/)
:::
:::

# Questions? {.center background-color="#1a3a2e"}

Slides, demo `.R` files, and live-coding companion at:

[**github.com/SiddharthaSBora/r-dataviz-webinar**](https://github.com/SiddharthaSBora/r-dataviz-webinar)
